Containers and R---the Rockerverse and beyond

### 2020-01-16 manuscript.Rmd
update conclusion, protected space for Docker Hub

The R community keeps growing. This can be seen in the number of new packages on CRAN, which keeps on growing exponentially \citep{cran:2019}, but also in the numbers of conferences, open educational resources, meetups, unconferences, and companies taking up, as exemplified by the useR! conference series\footnote{\href{https://www.r-project.org/conferences/}{https://www.r-project.org/conferences/}}, the global growth of the R and R-Ladies user groups\footnote{\href{https://www.r-consortium.org/blog/2019/09/09/r-community-explorer-r-user-groups}{https://www.r-consortium.org/blog/2019/09/09/r-community-explorer-r-user-groups}, \href{https://www.r-consortium.org/blog/2019/08/12/r-community-explorer}{https://www.r-consortium.org/blog/2019/08/12/r-community-explorer}}, or the foundation and impact of the R&nbsp;Consortium\footnote{\href{https://www.r-consortium.org/news/announcements}{https://www.r-consortium.org/news/announcements}, \href{https://www.r-consortium.org/blog/2019/11/14/data-driven-tracking-and-discovery-of-r-consortium-activities}{https://www.r-consortium.org/blog/2019/11/14/data-driven-tracking-and-discovery-of-r-consortium-activities}}.
All this cements the role of R as the _lingua&nbsp;franca_ of statistics, data visualisation, and computational research.
Coinciding with the rise of R was the advent of [Docker](https://en.wikipedia.org/wiki/Docker_(software)) as a general tool for distribution and deployment of server applications---in fact, Docker can be called the _lingua&nbsp;franca_ of describing computing environments and packaging software.
Combining both these topics, the _Rocker&nbsp;Project_ ([https://www.rocker-project.org/](https://www.rocker-project.org/)) provides images with R (see the next Section for more details).
The considerable uptake and continued evolution of the Rocker Project has lead to numerous projects extending or building upon Rocker images, ranging from reproducible research to production deployments.
This article presents this _Rockerverse_ of projects across all development stages: early demonstrations, working prototypes, and mature products.
We also introduce related activities connecting the R language and environment with other containerisation solutions.
The main contributions is a coherent picture of the current lay of the land of using containers in, with, and for R.

@@

**Microsoft R Open** ([MRO](https://github.com/nuest/mro-docker)) is an R distribution formerly known as Revolution R Open (RRO) before Revolution Analytics was acquired by Microsoft.
MRO is compatible with main R and its packages.
"It includes additional capabilities for improved performance, reproducibility, and platform support." \citep{microsoft_mro}
Most notably these capabilities are the MRAN repository, which is enabled by default, and the (optional) integration with [Intel® Math Kernel Library](https://software.intel.com/en-us/mkl) (MKL) for multi-threading in linear algebra operations \citep{microsoft_multithread}.
MRO does not provide official Docker images, but a set of community-maintained `Dockerfile`s and Docker images `nuest/mro` are available on Docker Hub at [https://github.com/nuest/mro-docker](https://github.com/nuest/mro-docker) and on GitHub at [https://github.com/nuest/mro-docker](https://github.com/nuest/mro-docker) respectively.
The images are inspired by the Rocker images and can be used much in the same fashion, effectively a drop-in replacement allowing users to quickly evaluate if the benefits of MRO + Intel® MKL apply to their use case.
Version-tagged images are provided for the latest bugfix release of recent R versions.
Extended license information about MKL is printed at every startup.

**Renjin** is an interpreter for the R language running on top of the [Java Virtual Machine](https://en.wikipedia.org/wiki/Java_virtual_machine) (JVM) providing full two-way access between Java and R code \citep{wikipedia_renjin_2018}.
It was developed to combine the benefits of R, such scripting and extension packages, with the JVM's advantages in the areas of security, cross-platform availability, and established position in enterprise settings.
R extension packages need to be specially compiled and are distributed via the Java package manager [Apache Maven](https://en.wikipedia.org/wiki/Apache_Maven), cf. [http://packages.renjin.org/packages](http://packages.renjin.org/packages) for available packages.
Packages are loaded on demand, i.e., at the first call to `library()`.
Not all R packages, especially one linking to binary libraries, are available, e.g., `rgdal` \footnote{\href{http://packages.renjin.org/package/org.renjin.cran/rgdal/1.4-4/build/1}{http://packages.renjin.org/package/org.renjin.cran/rgdal/1.4-4/build/1}}.
There are no offical Docker images for Renjin, but community-maintained images for selected releases only are available under `nuest/renjin` on Docker Hub and GitHub at [https://hub.docker.com/r/nuest/renjin](https://hub.docker.com/r/nuest/renjin) and [https://github.com/nuest/renjin-docker](https://github.com/nuest/renjin-docker) respectively.
These images expose the command line interface of Renjin in a similar fashion as Rocker images and allow an easy evaluation of Renjin's suitability, but are not intended for production use.

**[pqR](http://www.pqr-project.org/)** is a "a pretty quick version of R".
_pqR_ attempts to improve R on some opinionated issues in the R language and is the basis for experimental features, e.g., automatic diffentiation\footnote{\href{https://riotworkshop.github.io/abstracts/riot-2019-pqr.txt}{https://riotworkshop.github.io/abstracts/riot-2019-pqr.txt}}.
The source code development [on GitHub](https://github.com/radfordneal/pqR/) is a one man project and it does not provide any Docker images.
But especially disruptive approaches may contribute to the development of the R ecosystem, so the `nuest/pqr` images on Docker Hub and GitHub at [https://hub.docker.com/r/nuest/pqr/](https://hub.docker.com/r/nuest/pqr/) and [https://github.com/nuest/pqr-docker](https://github.com/nuest/pqr-docker) respectively.

**FastR** is "A high-performance implementation of the R programming language, built on GraalVM" (https://github.com/oracle/fastr).
It is developed by Oracle, connects R to the GraalVM ecosystem \citep{wikipedia_graalvm_2019}, and also claims superior performance but also targets full compatibility with base R\footnote{\href{https://github.com/oracle/fastr}{https://github.com/oracle/fastr}}.
There are no offical Docker images provided, but `nuest/fastr` images and `Dockerfile` are independently maintained on [Docker Hub](https://cloud.docker.com/repository/docker/nuest/fastr/) and [GitHub](https://github.com/nuest/fastr-docker) respectively.

@@

**rocker-win** ([https://github.com/nuest/rocker-win](https://github.com/nuest/rocker-win)) is a proof of concept for running R in Windows-based containers \citep{nust_rocker-win_2019}.
It provides selected R versions using three different base images using `microsoft/windowsservercore` and `mcr.microsoft.com/windows/servercore`.
These base images match different Windows versions: Windows Server 2019, Windows Server 2016, and Windows Server, version 1803.
The images are built using two CI services, [Travis CI](https://docs.travis-ci.com/user/reference/windows/) and [Appveyor](https://www.appveyor.com/docs/windows-images-software/), which provide the different Windows versions and therefore support different base images, and published automatically [on Docker Hub](https://hub.docker.com/r/nuest/rocker-win).
The images are built from manually maintained seperate `Dockerfile`s, are tagged with both Windows-variant and R version, e.g., `nuest/rocker-win:ltsc2019-3.6.2` or `nuest/rocker-win:1803-latest`, and run the `R.exe` process by default.
All these images can be run on Docker for Desktop on Windows 10.
New R versions are only added on demand.
The explorative rocker-win project demonstrates the variety and pontentially confusing complexity of Windows on Docker, e.g., because of the different base images, but also provides a starting point for any Windows-constrained user to leverage containerisation, including examples for apps using \pkg{plumber\, \pkg{Shiny}, or R packages with system dependencies such as \pkg{sf}.
The latter is quite comfortable and fast actually, compared to installation from source on Linux, as R ships pre-compiled binaries for Windows.
However, compared to the various Linux images, the footprint of Windows images is quite large.

@@

\label{rhub}
The **R-hub** project provides _"a collection of services to help R package development"_, with the package builder as the most prominent one\citep{r-hub_docs_2019}.
The builder allows R package developers to check their R package on different platforms and R versions using a web form or the package \pkg{rhub} \citep{csardi_rhub_2019}.
The builder uses Docker containers to conduct these checks, taking advantages of their well-defined environments and sandboxing.
The `Dockerfile`s and images are published on Docker Hub\footnote{\href{https://hub.docker.com/u/rhub}{https://hub.docker.com/u/rhub}} and GitHub\footnote{\href{https://github.com/r-hub/rhub-linux-builders}{https://github.com/r-hub/rhub-linux-builders}} respectively.
The images comprise release and development builds of R running in the base images of the [Debian](https://www.debian.org/), [Ubuntu](https://ubuntu.com/), [Fedora](https://getfedora.org/), and [CentOS](https://centos.org/) linux distributions (Arch Linux is under development\footnote{\href{https://github.com/r-hub/rhub-linux-builders/pull/41}{https://github.com/r-hub/rhub-linux-builders/pull/41}}, see `rhub::local_check_linux_images()` for a full list), and are intended for debugging (see Section&nbsp;\nameref{development}) not for reproducible workflows.
_The rationale_ of not re-using the Rocker image stack is the high number of operatings systems and configurations, which can be more uniformly maintained in a seperate suite of images.
Also, a common structure across the respective image stacks of the operating systems, e.g., for `debian`, adding GCC in `debian-gcc` and then R's development version in `debian-gcc-devel`.
The platform also has a specific user configuration differing from the setup covering typical applications of Rocker images, and the images include considerably more software, such as a fairly complete LaTeX installation, to provide an experience closer to CRAN.

@@

**Kaggle** provides the [`gcr.io/kaggle-images/rstats`](https://hub.docker.com/r/kaggle/rstats) image (previously `kaggle/rstats`) and [corresponding `Dockerfile`](https://github.com/Kaggle/docker-rstats) for usage in their Machine Learning competitions and easy access to the associated datasets.
It includes machine learling libraries such as Tensorflow and Keras, and also configures the \pkg{reticulate} package.
The image uses a base image with _all packages from CRAN_, `gcr.io/kaggle-images/rcran`, which requires a Google Cloud Build as Docker Hub would time out\footnote{Originally, a stacked collection of over 20 images with automated builds on Docker Hub was used, see \href{https://web.archive.org/web/20190606043353/http://blog.kaggle.com/2016/02/05/how-to-get-started-with-data-science-in-containers/}{https://web.archive.org/web/20190606043353/http://blog.kaggle.com/2016/02/05/how-to-get-started-with-data-science-in-containers/} and \href{https://hub.docker.com/r/kaggle/rcran/dockerfile}{https://hub.docker.com/r/kaggle/rcran/dockerfile}}.
The final extracted image size is over 25GB, which makes it debatable if having everything available is actually convenient.

@@

- [dynwrap](https://github.com/dynverse/dynwrap_containers/blob/master/.travis.yml) [@rcannood]
  - For this project, we use travis-ci to build rocker-derived containers, test them, and only push them to docker hub (from travis-ci.org) if the integration tests succeed.
- Google Cloud Build
  - https://cloud.google.com/cloud-build/
  - Cloud Build is another continuous intergation service that helps move the workload of building the Docker images to an online service.  Cloud Build can be set up to build the Dockerfiles on each GitHub commit or release.  This means you do not need to build the Docker images locally, which can tie up resources since Docker images can be several GBs and take a long time to compile. Google Cloud Build works alongside Google Container Registry to allow you to build private and public Docker images, which allows you to build up your own dependency graphs for downstream applications.
  - https://code.markedmondson.me/googleCloudRunner/articles/cloudbuild.html

@@

- Docker images for cloud services
  - A popular application of using Docker is to create the environments that can run in parallel for speeding up R code.  For example, [`googleComputeEngineR`](https://CRAN.R-project.org/package=googleComputeEngineR)'s `gce_vm_cluster()` function can create clusters of 2 or more virtual machines, running multi-CPU architectures.  Instead of running a local R script with the local CPU and RAM restrictions, the same code can be processed on all CPU threads of the cluster of machines in the cloud, all running in a Docker container with the same R environments.   This is achieved through `googleComputeEngineR`'s integration with the R parralisation library [`library(future)` by Henrik Bengtsson]( https://CRAN.R-project.org/package=future).  Local R computation can be thrown up to a multi-CPU and VM environment to achieve parrell computation in a few lines of R code in your local session. - [some demonstrations are available here](https://cloudyr.github.io/googleComputeEngineR/articles/massive-parallel.html). 
- [Google Cloud Run](https://cloud.run) - This is a CaaS (Containers as a Service) that lets you launch a Docker container without worrying about underlying infrastructure.  This dispenses with the developer creating a cloud server to run the Docker image on, by abstracting away those servers to a more serverless configuration.  Cloud Run lets you run your code on top of a managed or your own Kubernetes cluster running Knative, and can accept any Docker image.  The service takes care of network ingress, scaling machines up and down to zero, authentication and authorisation, all features which are non-trivial for a developer to create on their own.  This can be used to scale up R code to millions of instances if they need to, with little or no changes to existing code. An R implementation is shown here at [cloudRunR](https://github.com/MarkEdmondson1234/cloudRunR) which uses Cloud Run to create a scalable R plumber API.
  - How does this relate to https://code.markedmondson.me/googleCloudRunner/index.html ?`
- `batchtools` \citep{Lang2017batchtools} can [schedule jobs with Docker Swarm](https://mllg.github.io/batchtools/reference/makeClusterFunctionsDocker.html)
- scalable deployments, e.g., start with numerous Shiny talks mentioning Rocker at useR!2017
- [dynmethods](https://github.com/dynverse/dynmethods) [@rcannood]: In order to evaluate ±50 computational methods which all used different environments (R, Python, C++, ...), we wrapped each of them in a docker container and can execute these methods from R. Again, all of these containers are being built on travis-ci, and will only be pushed to docker hub if the integration test succeeds.
- \pkg{drake}, https://docs.ropensci.org/drake/index.html?q=docker#with-docker

@@

- researchcompendia.science
- https://github.com/benmarwick/rrtools

Containers provide a high degree of isolation that is often desirable when attempting to capture a specific computational environment so that others can reproduce and extend a research result. Many computationally intensive research projects depend on specific versions of original and third-party software packages in diverse languages, joined together to form a  pipeline through which data flows. New releases of even just a single piece of software in this pipeline can break the entire workflow, making it difficult for others to use, and making the original the results irreproducible. The chance of a substantial disruption like this is high in a multi-year research project where key pieces of third-party software may have several major updates over the duration of the project. Containers give the researcher an isolated environment to assemble these research pipelines with specific versions of software. Research workflows in containers are safe from contamination from other activities occuring on the researcher's computer, for example the installation of newest version of packages for teaching demonstrations.

In many fields we see that a typical unit of research for a container is a research report or journal article, where the container holds the compendium, or self-contained set of data (or connections to data elsewhere) and code files needed to fully reproduce the article [@doi:10.1080/00031305.2017.1375986]. Images of compendium containers can be hosted on services such as Docker Hub for convienient sharing among collaborators and others. Similarly, packages such as containerit and dockerfiler can be used to create a Dockerfile to be archived with a compendium on a data repository (e.g. Zenodo, Dataverse, Figshare, OSF). A typial compendium Dockerfile will pull a rocker image fixed to a specific version of R, and install R packages from the MRAN repository to ensure the package versions are tied to a specific date, rather than the most recent version. A future researchers can download then compendium from the repository and run the included Dockerfile build a new image that recreates the computational environment used to produce the original research results.

- \pkg{renv}, https://rstudio.github.io/renv/articles/docker.html
- what to copy in and what not

@@

Third, Docker images can be used **on CI platforms** to streamline the testing of packages.
\citet{ye_docker_2019} describes how they speed up the process of testing by running tasks on [Travis&nbsp;CI](https://travis-ci.org/) within a container using `docker exec`, e.g., the package check or rendering of documentation.
\citet{cardozo_faster_2018}, also on Travis&nbsp;CI saved time by re-using the testing image as the base iamge for an image intended for publication on Docker Hub.
Especially for long running tests or complex system dependencies, these approaches seems to improve the development.
Not due to a concern about time, but to control the environment used on a CI server, even this manuscript was rendered into a PDF and deployed to a GitHub-hosted website (see `.travis.yml` and `Dockerfile` in the manuscript repository).
This gives on the one hand easy access after every update of the R Markdown source code, and on the other hand a second controlled environment making sure that the article renders successfully and correctly.

@@

The R packages and use cases presented show the positive drive in the community to innovate, but also the challenge of keeping up with a continuously evolving landscape.
The use cases contributed by co-authors also have a degree of overlap.
A common language and understanding of good practices is still taking shape.
The easiness with which one can create a complex software systems, such as an independent Docker image stack, to serve one's specific needs currently leads to some fragmentation.
It is unclear if Docker alone as a core technology is really a feasible level of collaboration and abstraction.
Furthermore fragmentation may not be a bad sign but instead a reflection of a growing market, which is able to sustain multiple related efforts.
With the maturing of core building blocks, such as the Rocker suite of images, more systems will be built successfully but will also be behind the curtains.

However, at least _on the level of R packages_, some consolidation seems in order, e.g., to reduce the number of packages creating `Dockerfile`s from R code or controlling the Docker daemon with R code.
It remains to be seen which approach to control Docker, via the Docker API as `stevedore` or via system calls as `dockyard`/`docker`/`dockr`, is more sustainable, or if the question will be answered by the endurance of maintainers and sufficient funding.
Similarly, the capturing of environments and their serialization in form of a `Dockerfile` currently happens at different levels of abstraction and re-use of functionality seems reasonable, e.g. `liftr` could generate the environment with `containerit`, which in turn may use `dockerfiler` for low level objects representing a `Dockerfile`.
In this consolidation, the Rocker Project could play the role of a unifying and coordinating entity in the future.
Though for the moment, the sign of the times points to more experimentation and feature growth, e.g. images for GPU-based computing and artificial intelligence.
Even with coding being more and more accepted as a required, and achievable skill, an easier access, for example by exposing containerisation benefits via simple user interfaces in the users' IDE, could be an important next step.
Currently containerisation happens more in the background at the system level.

>>>>

The R community keeps growing. This can be seen in the number of new packages on CRAN, which keeps on growing exponentially \citep{cran:2019}, but also in the numbers of conferences, open educational resources, meetups, unconferences, and companies taking up, as exemplified by the useR! conference series\footnote{\href{https://www.r-project.org/conferences/}{https://www.r-project.org/conferences/}}, the global growth of the R and R-Ladies user groups\footnote{\href{https://www.r-consortium.org/blog/2019/09/09/r-community-explorer-r-user-groups}{https://www.r-consortium.org/blog/2019/09/09/r-community-explorer-r-user-groups}, \href{https://www.r-consortium.org/blog/2019/08/12/r-community-explorer}{https://www.r-consortium.org/blog/2019/08/12/r-community-explorer}}, or the foundation and impact of the R&nbsp;Consortium\footnote{\href{https://www.r-consortium.org/news/announcements}{https://www.r-consortium.org/news/announcements}, \href{https://www.r-consortium.org/blog/2019/11/14/data-driven-tracking-and-discovery-of-r-consortium-activities}{https://www.r-consortium.org/blog/2019/11/14/data-driven-tracking-and-discovery-of-r-consortium-activities}}.
All this cements the role of R as the _lingua&nbsp;franca_ of statistics, data visualisation, and computational research.
Coinciding with the rise of R was the advent of [Docker](https://en.wikipedia.org/wiki/Docker_(software)) as a general tool for distribution and deployment of server applications---in fact, Docker can be called the _lingua&nbsp;franca_ of describing computing environments and packaging software.
Combining both these topics, the _Rocker&nbsp;Project_ ([https://www.rocker-project.org/](https://www.rocker-project.org/)) provides images with R (see the next Section for more details).
The considerable uptake and continued evolution of the Rocker Project has lead to numerous projects extending or building upon Rocker images, ranging from reproducible research\footnote{"Reproducible" in the sense of the _Claerbout/Donoho/Peng_ terminology \citep{barba_terminologies_2018}.} to production deployments.
This article presents this _Rockerverse_ of projects across all development stages: early demonstrations, working prototypes, and mature products.
We also introduce related activities connecting the R language and environment with other containerisation solutions.
The main contributions is a coherent picture of the current lay of the land of using containers in, with, and for R.

@@

**Microsoft R Open** ([MRO](https://github.com/nuest/mro-docker)) is an R distribution formerly known as Revolution R Open (RRO) before Revolution Analytics was acquired by Microsoft.
MRO is compatible with main R and its packages.
"It includes additional capabilities for improved performance, reproducibility, and platform support." \citep{microsoft_mro}
Most notably these capabilities are the MRAN repository, which is enabled by default, and the (optional) integration with [Intel® Math Kernel Library](https://software.intel.com/en-us/mkl) (MKL) for multi-threading in linear algebra operations \citep{microsoft_multithread}.
MRO does not provide official Docker images, but a set of community-maintained `Dockerfile`s and Docker images `nuest/mro` are available on Docker&nbsp;Hub at [https://github.com/nuest/mro-docker](https://github.com/nuest/mro-docker) and on GitHub at [https://github.com/nuest/mro-docker](https://github.com/nuest/mro-docker) respectively.
The images are inspired by the Rocker images and can be used much in the same fashion, effectively a drop-in replacement allowing users to quickly evaluate if the benefits of MRO + Intel® MKL apply to their use case.
Version-tagged images are provided for the latest bugfix release of recent R versions.
Extended license information about MKL is printed at every startup.

**Renjin** is an interpreter for the R language running on top of the [Java Virtual Machine](https://en.wikipedia.org/wiki/Java_virtual_machine) (JVM) providing full two-way access between Java and R code \citep{wikipedia_renjin_2018}.
It was developed to combine the benefits of R, such scripting and extension packages, with the JVM's advantages in the areas of security, cross-platform availability, and established position in enterprise settings.
R extension packages need to be specially compiled and are distributed via the Java package manager [Apache Maven](https://en.wikipedia.org/wiki/Apache_Maven), cf. [http://packages.renjin.org/packages](http://packages.renjin.org/packages) for available packages.
Packages are loaded on demand, i.e., at the first call to `library()`.
Not all R packages, especially one linking to binary libraries, are available, e.g., `rgdal` \footnote{\href{http://packages.renjin.org/package/org.renjin.cran/rgdal/1.4-4/build/1}{http://packages.renjin.org/package/org.renjin.cran/rgdal/1.4-4/build/1}}.
There are no offical Docker images for Renjin, but community-maintained images for selected releases only are available under `nuest/renjin` on Docker&nbsp;Hub and GitHub at [https://hub.docker.com/r/nuest/renjin](https://hub.docker.com/r/nuest/renjin) and [https://github.com/nuest/renjin-docker](https://github.com/nuest/renjin-docker) respectively.
These images expose the command line interface of Renjin in a similar fashion as Rocker images and allow an easy evaluation of Renjin's suitability, but are not intended for production use.

**[pqR](http://www.pqr-project.org/)** is a "a pretty quick version of R".
_pqR_ attempts to improve R on some opinionated issues in the R language and is the basis for experimental features, e.g., automatic diffentiation\footnote{\href{https://riotworkshop.github.io/abstracts/riot-2019-pqr.txt}{https://riotworkshop.github.io/abstracts/riot-2019-pqr.txt}}.
The source code development [on GitHub](https://github.com/radfordneal/pqR/) is a one man project and it does not provide any Docker images.
But especially disruptive approaches may contribute to the development of the R ecosystem, so the `nuest/pqr` images on Docker&nbsp;Hub and GitHub at [https://hub.docker.com/r/nuest/pqr/](https://hub.docker.com/r/nuest/pqr/) and [https://github.com/nuest/pqr-docker](https://github.com/nuest/pqr-docker) respectively.

**FastR** is "A high-performance implementation of the R programming language, built on GraalVM" (https://github.com/oracle/fastr).
It is developed by Oracle, connects R to the GraalVM ecosystem \citep{wikipedia_graalvm_2019}, and also claims superior performance but also targets full compatibility with base R\footnote{\href{https://github.com/oracle/fastr}{https://github.com/oracle/fastr}}.
There are no offical Docker images provided, but `nuest/fastr` images and `Dockerfile` are independently maintained on [Docker&nbsp;Hub](https://cloud.docker.com/repository/docker/nuest/fastr/) and [GitHub](https://github.com/nuest/fastr-docker) respectively.

@@

**rocker-win** ([https://github.com/nuest/rocker-win](https://github.com/nuest/rocker-win)) is a proof of concept for running R in Windows-based containers \citep{nust_rocker-win_2019}.
It provides selected R versions using three different base images using `microsoft/windowsservercore` and `mcr.microsoft.com/windows/servercore`.
These base images match different Windows versions: Windows Server 2019, Windows Server 2016, and Windows Server, version 1803.
The images are built using two CI services, [Travis CI](https://docs.travis-ci.com/user/reference/windows/) and [Appveyor](https://www.appveyor.com/docs/windows-images-software/), which provide the different Windows versions and therefore support different base images, and published automatically [on Docker&nbsp;Hub](https://hub.docker.com/r/nuest/rocker-win).
The images are built from manually maintained seperate `Dockerfile`s, are tagged with both Windows-variant and R version, e.g., `nuest/rocker-win:ltsc2019-3.6.2` or `nuest/rocker-win:1803-latest`, and run the `R.exe` process by default.
All these images can be run on Docker for Desktop on Windows 10.
New R versions are only added on demand.
The explorative rocker-win project demonstrates the variety and pontentially confusing complexity of Windows on Docker, e.g., because of the different base images, but also provides a starting point for any Windows-constrained user to leverage containerisation, including examples for apps using \pkg{plumber\, \pkg{Shiny}, or R packages with system dependencies such as \pkg{sf}.
The latter is quite comfortable and fast actually, compared to installation from source on Linux, as R ships pre-compiled binaries for Windows.
However, compared to the various Linux images, the footprint of Windows images is quite large.

@@

\label{rhub}
The **R-hub** project provides _"a collection of services to help R package development"_, with the package builder as the most prominent one\citep{r-hub_docs_2019}.
The builder allows R package developers to check their R package on different platforms and R versions using a web form or the package \pkg{rhub} \citep{csardi_rhub_2019}.
The builder uses Docker containers to conduct these checks, taking advantages of their well-defined environments and sandboxing.
The `Dockerfile`s and images are published on Docker&nbsp;Hub\footnote{\href{https://hub.docker.com/u/rhub}{https://hub.docker.com/u/rhub}} and GitHub\footnote{\href{https://github.com/r-hub/rhub-linux-builders}{https://github.com/r-hub/rhub-linux-builders}} respectively.
The images comprise release and development builds of R running in the base images of the [Debian](https://www.debian.org/), [Ubuntu](https://ubuntu.com/), [Fedora](https://getfedora.org/), and [CentOS](https://centos.org/) linux distributions (Arch Linux is under development\footnote{\href{https://github.com/r-hub/rhub-linux-builders/pull/41}{https://github.com/r-hub/rhub-linux-builders/pull/41}}, see `rhub::local_check_linux_images()` for a full list), and are intended for debugging (see Section&nbsp;\nameref{development}) not for reproducible workflows.
_The rationale_ of not re-using the Rocker image stack is the high number of operatings systems and configurations, which can be more uniformly maintained in a seperate suite of images.
Also, a common structure across the respective image stacks of the operating systems, e.g., for `debian`, adding GCC in `debian-gcc` and then R's development version in `debian-gcc-devel`.
The platform also has a specific user configuration differing from the setup covering typical applications of Rocker images, and the images include considerably more software, such as a fairly complete LaTeX installation, to provide an experience closer to CRAN.

@@

**Kaggle** provides the [`gcr.io/kaggle-images/rstats`](https://hub.docker.com/r/kaggle/rstats) image (previously `kaggle/rstats`) and [corresponding `Dockerfile`](https://github.com/Kaggle/docker-rstats) for usage in their Machine Learning competitions and easy access to the associated datasets.
It includes machine learling libraries such as Tensorflow and Keras, and also configures the \pkg{reticulate} package.
The image uses a base image with _all packages from CRAN_, `gcr.io/kaggle-images/rcran`, which requires a Google Cloud Build as Docker&nbsp;Hub would time out\footnote{Originally, a stacked collection of over 20 images with automated builds on Docker~Hub was used, see \href{https://web.archive.org/web/20190606043353/http://blog.kaggle.com/2016/02/05/how-to-get-started-with-data-science-in-containers/}{https://web.archive.org/web/20190606043353/http://blog.kaggle.com/2016/02/05/how-to-get-started-with-data-science-in-containers/} and \href{https://hub.docker.com/r/kaggle/rcran/dockerfile}{https://hub.docker.com/r/kaggle/rcran/dockerfile}}.
The final extracted image size is over 25GB, which makes it debatable if having everything available is actually convenient.

@@

- [dynwrap](https://github.com/dynverse/dynwrap_containers/blob/master/.travis.yml) [@rcannood]
  - For this project, we use travis-ci to build rocker-derived containers, test them, and only push them to Docker&nbsp;Hub (from travis-ci.org) if the integration tests succeed.
- Google Cloud Build
  - https://cloud.google.com/cloud-build/
  - Cloud Build is another continuous intergation service that helps move the workload of building the Docker images to an online service.  Cloud Build can be set up to build the Dockerfiles on each GitHub commit or release.  This means you do not need to build the Docker images locally, which can tie up resources since Docker images can be several GBs and take a long time to compile. Google Cloud Build works alongside Google Container Registry to allow you to build private and public Docker images, which allows you to build up your own dependency graphs for downstream applications.
  - https://code.markedmondson.me/googleCloudRunner/articles/cloudbuild.html

@@

- Docker images for cloud services
  - A popular application of using Docker is to create the environments that can run in parallel for speeding up R code.  For example, [`googleComputeEngineR`](https://CRAN.R-project.org/package=googleComputeEngineR)'s `gce_vm_cluster()` function can create clusters of 2 or more virtual machines, running multi-CPU architectures.  Instead of running a local R script with the local CPU and RAM restrictions, the same code can be processed on all CPU threads of the cluster of machines in the cloud, all running in a Docker container with the same R environments.   This is achieved through `googleComputeEngineR`'s integration with the R parralisation library [`library(future)` by Henrik Bengtsson]( https://CRAN.R-project.org/package=future).  Local R computation can be thrown up to a multi-CPU and VM environment to achieve parrell computation in a few lines of R code in your local session. - [some demonstrations are available here](https://cloudyr.github.io/googleComputeEngineR/articles/massive-parallel.html). 
- [Google Cloud Run](https://cloud.run) - This is a CaaS (Containers as a Service) that lets you launch a Docker container without worrying about underlying infrastructure.  This dispenses with the developer creating a cloud server to run the Docker image on, by abstracting away those servers to a more serverless configuration.  Cloud Run lets you run your code on top of a managed or your own Kubernetes cluster running Knative, and can accept any Docker image.  The service takes care of network ingress, scaling machines up and down to zero, authentication and authorisation, all features which are non-trivial for a developer to create on their own.  This can be used to scale up R code to millions of instances if they need to, with little or no changes to existing code. An R implementation is shown here at [cloudRunR](https://github.com/MarkEdmondson1234/cloudRunR) which uses Cloud Run to create a scalable R plumber API.
  - How does this relate to https://code.markedmondson.me/googleCloudRunner/index.html ?`
- `batchtools` \citep{Lang2017batchtools} can [schedule jobs with Docker Swarm](https://mllg.github.io/batchtools/reference/makeClusterFunctionsDocker.html)
- scalable deployments, e.g., start with numerous Shiny talks mentioning Rocker at useR!2017
- [dynmethods](https://github.com/dynverse/dynmethods) [@rcannood]: In order to evaluate ±50 computational methods which all used different environments (R, Python, C++, ...), we wrapped each of them in a docker container and can execute these methods from R. Again, all of these containers are being built on travis-ci, and will only be pushed to Docker&nbsp;Hub if the integration test succeeds.
- \pkg{drake}, https://docs.ropensci.org/drake/index.html?q=docker#with-docker

@@

- researchcompendia.science
- https://github.com/benmarwick/rrtools
- \pkg{renv}, https://rstudio.github.io/renv/articles/docker.html
- what to copy in and what not

@@

Third, Docker images can be used **on CI platforms** to streamline the testing of packages.
\citet{ye_docker_2019} describes how they speed up the process of testing by running tasks on [Travis&nbsp;CI](https://travis-ci.org/) within a container using `docker exec`, e.g., the package check or rendering of documentation.
\citet{cardozo_faster_2018}, also on Travis&nbsp;CI saved time by re-using the testing image as the base iamge for an image intended for publication on Docker&nbsp;Hub.
Especially for long running tests or complex system dependencies, these approaches seems to improve the development.
Not due to a concern about time, but to control the environment used on a CI server, even this manuscript was rendered into a PDF and deployed to a GitHub-hosted website (see `.travis.yml` and `Dockerfile` in the manuscript repository).
This gives on the one hand easy access after every update of the R Markdown source code, and on the other hand a second controlled environment making sure that the article renders successfully and correctly.

@@

The R packages and use cases presented show the positive drive in the community to innovate, but also the challenge of keeping up with a continuously evolving landscape.
The use cases contributed by co-authors also have a degree of overlap, which can be expected as a common language and understanding of good practices is still taking shape.
Also, the ease with which one can create a complex software systems, such as an independent Docker image stack, to serve one's specific needs leads to some fragmentation.
Furthermore fragmentation may not be a bad sign but instead a reflection of a growing market, which is able to sustain multiple related efforts.
With the maturing of core building blocks, such as the Rocker suite of images, more systems will be built successfully but will also be behind the curtains.
Docker alone, as a flexible core technology, is not a feasible level of collaboration and abstraction.
Instead, the use cases and applications observed in this work provide a more useful division.

Still, at least _on the level of R packages_ some consolidation seems in order, e.g., to reduce the number of packages creating `Dockerfile`s from R code or controlling the Docker daemon with R code.
It remains to be seen which approach to control Docker, via the Docker API as `stevedore` or via system calls as `dockyard`/`docker`/`dockr`, is more sustainable, or if the question will be answered by the endurance of maintainers and sufficient funding.
Similarly, the capturing of environments and their serialization in form of a `Dockerfile` currently happens at different levels of abstraction and re-use of functionality seems reasonable, e.g. `liftr` could generate the environment with `containerit`, which in turn may use `dockerfiler` for low level objects representing a `Dockerfile`.
In this consolidation, the Rocker Project could play the role of a unifying and coordinating entity in the future.
Though for the moment, the sign of the times points to more experimentation and feature growth, e.g. images for GPU-based computing and artificial intelligence.
Even with coding being more and more accepted as a required, and achievable skill, an easier access, for example by exposing containerisation benefits via simple user interfaces in the users' IDE, could be an important next step.
Currently containerisation happens more in the background at the system level.

New features, which make complex workflows accessible and reproducible, and variety in packages, even when they have overlapping features, are a signal and a support for a growing user base.
This growth is possibly the most important goal for the foreseeable future in the _Rockerverse_, and just like the Rocker images have matured over years of use and millions of runs the new ideas and prototypes will have to proof themselves.
It should be noted that the dominant position of Docker is a blessing and a curse for these goals.
It could be wise to start experimenting with non-Docker containerisation tools now, e.g. R packages interfacing with other container engines such as [podman/buildah](https://github.com/containers/libpod) or [rkt](https://coreos.com/rkt/), or an R package for creating Singularity image recipe files.
Such efforts help to avoid lock-in and to design sustainable workflows based on concepts of _containerisation_, not on their implementation in Docker.
If adoption of containerisation and R continue to grow, the missing pieces for a success predominantly lie in (a) coordination and documentation of activities to reduce repeated work in favour of open collaboration, (b) the sharing of lessons learned from use cases to build common knowledge and language, and (c) a sustainable continuation and funding for all of development, community support, and education.
A concrete effort to work towards these pieces is to sustain the structure and captured status quo from this work in form of a _CRAN Task View on containerization_.
